Rice yield improvements by plant breeding are offset by decreases over time

### 2017-04-30 output/breed_paper.Rnw
Adjusts figure 1

\title{Rice yield improvement by plant breeding are offset by
  decreases over time}
\author[ucd]{Matthew B. Espe\corref{corresauthor}}
\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}

@@

%\linenumbers

@@

@@

Constraints on ariable land are increasing simultaneous with the need
to increase total production to meet growing demand, which has
necessitated harvesting more grain per unit land area. Historically,
agricultural research has been successful in staving off the Mathesian
disaster through continued yield improvement of staple crops, however
more and more production systems are experiencing plateaus in grain
yield \citep{Peng2010}. If continued improvements per unit land area
are no longer possible, increasing the area under cultivation will be
needed to meet demand, which carries with it undesirable ecological
implications. Therefore, it is of critical importance to better
understand why grain yields have not continued to increase.

CA rice production represents one such rice production system. Rice
yields in CA during the period from 1920 -- 1990 increased rapidly, but
since 1990 have increased only slowly if at all (Fig. 1
\ref{fig:avg_yields}). However, continued improvement of grain yield
as the result of plant breeding has been demonstrated since 1990
\citep{rrb}, with each new cultivar having a yield advantage over the
previous cultivars. Yet, these same reports show stagnant yields; for
example, at the year of release (1984), cultivar 'M-202' averaged
9,600 lbs acre\textsuperscript{-1}. In 2015, the yield of 'M-209' was
between 9,000 -- 10,000 lbs acre\textsuperscript{-1}. We sought to better
understand how grain yields have increased in the breeders' fields
without a corresponding increase in yield in the farmers' fields. Two
competing hypotheses were tested: (1) Grain yields have not improved
since 1990, or (2) grain yield improvements have been offset by an
associated decrease in yield once a cultivar was selected.

\begin{figure}[h]
<<fig=TRUE, echo=FALSE>>=
CA_data = read.csv("../data/CA_rice_yield.csv", stringsAsFactors = FALSE)
CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))
plot(Value * 1.12 ~ Year, CA_data,
     type = "l",
     xlab = "Year",
     ylab = expression("Average grain yield (kg ha"^-1~")"))

@
\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
  -- \Sexpr{max(CA_data$Year)} for California, USA. Rapid gains in
    grain yield were experienced from approximately 1950 -- 1990, but
    annual gains since 1990 have been slower.}
\label{fig:state_yield}
\end{figure}

@@

@@

res = mod_data$site == "RES"
plot_data = as.data.frame(tapply(mod_data$yield_lb[res] * 1.12,
                                 mod_data$id[res], mean))
plot_data$se = tapply(mod_data$yield_lb[res] * 1.12,
                      mod_data$id[res],
                      function(x)
                          sd(x)/sqrt(length(x)))

plot_data$tmp = tapply(mod_data$release_yr[res],
                       mod_data$id[res], unique)
colnames(plot_data) = c("yield_kg", "yield_se", "release_yr")

plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8000, 13000),
     xlim = c(1980,2019))
abline(a = (((fixef(RES_rel_fit)[1] * 1000) + 9000) * 1.12) -
           (2000 * fixef(RES_rel_fit)[2] * 1000),
       b = fixef(RES_rel_fit)[2] * 1000, col = "grey")
with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr, y = plot_data$yield_kg,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.8)

# ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
# polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
# text(0, 7000, bquote(.(round(fixef(RES_yrs_fit)[2] * 1000, 1))~lbs~year^-1))

# plot(yield_lb ~ release_yr, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
#      xlab = "Year of Release",
#      ylab =  expression("Yield ( lbs acre"^-1~")"))
# lines(x = yrs, y = ints[2,], lty = 2) 
# polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
text(1990, 8000, bquote(.(round(fixef(release_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))
@

@@

RES = res = mod_data$site == "RES"
plot_data = as.data.frame(tapply(mod_data$yield_lb[res] * 1.12,
                                 mod_data$yrs_in_trial[res], mean))
plot_data$se = tapply(mod_data$yield_lb[res] * 1.12,
                      mod_data$yrs_in_trial[res],
                      function(x)
                          sd(x)/sqrt(length(x)))

@@

plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Year since entry",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(7000, 13000),
     xlim = c(-5,33))
with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
abline(a = ((fixef(RES_yrs_fit)[1] * 1000) + 9000) * 1.12,
       b = fixef(RES_yrs_fit)[2] * 1000)
ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
#polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
text(0, 7000, bquote(.(round(fixef(RES_yrs_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))

@
\caption{The relationship between yield and the years since entry for
  medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. RES is considered
  the best managed site, and has the most plots per year of any of the
  sites. The estimated decrease over time is estimated to be stronger
  using only these data (\Sexpr{round(fixef(RES_yrs_fit)[2] * 1000, 1)}
  lbs year\textsuperscript{-1}), primarly due to lower variability.}
\label{fig:RES_yrs}
\end{figure}

\begin{figure}
<<fig=TRUE, fig.width = 10, fig.height = 5, out.width = '80%', echo=FALSE>>=
tars <- c("M202","M206")
par(mfrow = c(1, length(tars)),
    mar = c(4,0,2,0),
    oma = c(4,4,0.5,0.5))

invisible(sapply(tars, function(x){
    plot(yield_lb * 1.12 ~ year,
         data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
         xlim = c(1995, 2016), ylim = c(5000,14000),
         pch = 16, main = gsub("M", "M-", x),
         ylab = "",
         yaxt = "n",
         xlab = "Year")
    if(x == "M202"){
        axis(2)
        mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }
    # abline(lm(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
    abline(h = mean(mod_data$yield_lb[(mod_data$id == x & mod_data$site == "RES")] * 1.12),
           lty = 2)
}))

@@

<<fig=TRUE, echo=FALSE>>=
par(mfrow = c(1, length(tars)),
    mar = c(4,0,2,0),
    oma = c(4,4,0.5,0.5))

invisible(sapply(tars, function(x){
    plot(yield_lb*1.12 ~ yrs_in_trial,
         data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
         xlim = c(-6, 36), ylim = c(5000,14000),
         pch = 16, main = gsub("M","M-", x),
         ylab = "",
         yaxt = "n",
         xlab = "Years after release")
    if(x == "M202"){
        axis(2)
        mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }
    abline(lm(yield_lb*1.12 ~ yrs_in_trial,
              data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
}))

@@

\begin{itemize}
  \item H1: There is evidence of a small decrease in yield once a cultivar is
    released. This complicates comparisons between new entries and
    checks, since the check will decrease over time. This decrease
    will manifest as an apparent larger yield advantage of the new
    entry over the check.
  \item H2: There is evidence of a small increase in yield with each
    new released cultivar. Therefore, it appears that successive
    cultivar releases are not merely recovering to the old benchmark
    but are surpassing it. However, progress is slower than previously
    estimated (\Sexpr{round(fixef(release_fit)[2] * 1000, 1)}
    lbs acre\textsuperscript{-1} year\textsuperscript{-1} increase
    verses a previous estimate of 54 lbs acre\textsuperscript{-1}
    year\textsuperscript{-1}).
\end{itemize}

The practical impact of these results is:

\begin{enumerate}
\item Breeders need to be aware of the potential for the yield of
  checks to decrease over time. Due to this, comparisons to older checks
  might give a false sense of improvement. In order to have greater
  improvements in yield, the target yield improvement for a new cultivar
  needs to be higher to offset the impact of yield decrease once a
  cultivar is released.
\item Breeding is essential to maintaining yield in rice. Yield is not
  stable and without continuous effort, yield potential will decrease
  over time.

\end{enumerate}

\section{Supplemental figures}

\section{References}\label{references}

>>>>

\title{Rice yield improvements by plant breeding are offset by
  decreases over time}
\author[ucd]{Matthew B. Espe\corref{corresauthor}}
\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}

@@

\linenumbers

@@

prep_plot_data = function(x, idx)
{
    as.data.frame(do.call(rbind, tapply(x, idx, function(x){
        c(mean = mean(x), se = (sd(x)/sqrt(length(x))))
    })))
}

@@

Constraints on ariable land are increasing simultaneous with the need
to increase total production to meet growing demand \citep{Foley2011a,
  Godfray2010, Mueller2012}, which has
necessitated harvesting more grain per unit land area
\citep{Lobell2009, Tittonell2014}. Historically,
agricultural research has been successful in staving off the Mathesian
disaster through continued yield improvement of staple crops, however
more and more production systems are experiencing plateaus in grain
yield \citep{Grassini2013, Peng2010}. If continued improvements per unit land area
are no longer possible, increasing the area under cultivation will be
needed to meet demand, which carries with it undesirable ecological
implications. Therefore, it is of critical importance to better
understand why grain yields have not continued to increase in
instensified production systems.

California (CA) rice production represents one such rice production
system. Rice yields in CA during the period from 1920 -- 1990
increased rapidly, but since 1990 have increased more slowly (Fig. 1
\ref{fig:state_yield}). However, continued improvement of grain yield
as the result of plant breeding has been demonstrated since 1990
\citep{rrb}, with each new cultivar having a yield advantage over the
previous cultivars. Yet, these same reports show stagnant yields; for
example, at the year of release (1985), cultivar 'M-202' averaged
11,060 kg ha\textsuperscript{-1} \citep[14\% MC;][]{m202}. In 2015,
the average yield of the newest released cultivar 'M-209' was 10,841
kg ha\textsuperscript{-1} \citep{m209}. This suggests little to no
progress via breeding, however 'M-209' demonstrated an overall 10\%
yield advantage over 'M-202' in side-by-side comparisons \citep{m209}. We sought to
better understand how grain yields have increased in the breeders'
fields without a corresponding increase in yield in the farmers'
fields. Two competing hypotheses were tested: (1) Grain yields have
not improved since 1990, or (2) grain yield improvements have been
offset by an associated decrease in yield once a cultivar was
selected.

\begin{figure}[ht]
<<fig=TRUE, echo=FALSE>>=
CA_data = read.csv("../data/CA_rice_yield.csv", stringsAsFactors = FALSE)
CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))
CA_data = CA_data[order(CA_data$Year),]
plot(Value * 1.12 ~ Year, CA_data,
     # type = "l",
     pch = 16,
     xlab = "Year",
     ylab = expression("Average grain yield ( kg ha"^-1~")"))
mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
                  CA_data$Value * 1.12,
                  rep(mean(CA_data$Value[(nrow(CA_data) - 10):nrow(CA_data)] * 1.12), 5)),
                rep(1/11,11))
lines(x = c(1907:1911, CA_data$Year, 2017:2021), y = mv_avg)

@
\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
  -- \Sexpr{max(CA_data$Year)} for California, USA. Rapid gains in
    grain yield were experienced from approximately 1950 -- 1990, but
    annual gains since 1990 have been slower. The line represents the
    10 year moving average.}
\label{fig:state_yield}
\end{figure}

@@

The Rice Experiment Station (RES) of the California Cooperative Rice
Research Foundation

@@

res = mod_data$site == "RES"
plot_data = prep_plot_data(mod_data$yield_lb[res] * 1.12,
                           mod_data$id[res])

plot_data$tmp = tapply(mod_data$release_yr[res],
                       mod_data$id[res], unique)

colnames(plot_data) = c("yield_kg", "yield_se", "release_yr")

plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8000, 13000),
     xlim = c(1980,2019))
abline(a = (((fixef(RES_rel_fit)[1] * 1000) + 9000) * 1.12) -
           (2000 * fixef(RES_rel_fit)[2] * 1000),
       b = fixef(RES_rel_fit)[2] * 1000, col = "grey50")
with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr, y = plot_data$yield_kg,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.8)

text(1990, 8000, bquote(.(round(fixef(release_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))

@@

RES = res = mod_data$site == "RES"
plot_data = prep_plot_data(mod_data$yield_lb[res] * 1.12,
                           mod_data$yrs_in_trial[res])

@@

plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Year since entry",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(7000, 13000))
with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
abline(a = ((fixef(RES_yrs_fit)[1] * 1000) + 9000) * 1.12,
       b = fixef(RES_yrs_fit)[2] * 1000, col = "grey50")
ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
text(5, 7000, bquote(.(round(fixef(RES_yrs_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))

@
\caption{The relationship between yield and the years since entry for
  medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. RES is considered
  the best managed site, and has the most plots per year of any of the
  sites. The estimated decrease over time is estimated to be stronger
  using only these data (\Sexpr{round(fixef(RES_yrs_fit)[2] * 1000, 1)}
  lbs year\textsuperscript{-1}), primarly due to lower variability.}
\label{fig:RES_year}
\end{figure}

\begin{figure}[ht]
<<fig=TRUE, width = 8, height = 5, out.width = '85%', echo=FALSE>>=
tars <- c("M202","M206")
par(mfrow = c(1, length(tars)),
    mar = c(4,0,2,0),
    oma = c(4,4,0.5,0.5))

plot_data = lapply(tars, function(x) {
    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
    tmp = prep_plot_data(dd$yield_lb * 1.12, dd$year)
    tmp$year = as.numeric(row.names(tmp))
    tmp
    })

invisible(sapply(1:2, function(i){
    plot(mean ~ year,
         data = plot_data[[i]],
         xlim = c(1995, 2016), ylim = c(6500,14500),
         pch = 16, main = gsub("M","M-", tars[i]),
         ylab = "",
         yaxt = "n",
         xlab = "Years after release")
    segments(x0 = plot_data[[i]]$year,
             y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
             y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
    if(i == 1){
        axis(2)
        mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }
#     abline(lm(yield_lb*1.12 ~ year,
    #               data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
    # lty = 2)

    abline(h = mean(mod_data$yield_lb[(mod_data$id == tars[i] &
                                       mod_data$site == "RES")] * 1.12),
           lty = 1, col = "grey50")

}))

# invisible(sapply(tars, function(x){
#     plot(yield_lb * 1.12 ~ year,
#          data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
#          xlim = c(1995, 2016), ylim = c(6500,14500),
#          pch = 16,
#          col = rgb(0,0,0,0.5),
#          main = gsub("M", "M-", x),
#          ylab = "",
#          yaxt = "n",
#          xlab = "Year")
#     if(x == "M202"){
#         axis(2)
#         mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
#     }
#     # abline(lm(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
#     abline(h = mean(mod_data$yield_lb[(mod_data$id == x & mod_data$site == "RES")] * 1.12),
#            lty = 2)
# }))

@@

<<fig=TRUE, echo=FALSE, width = 8, height = 5>>=
par(mfrow = c(1, length(tars)),
    mar = c(4,0,2,0),
    oma = c(4,4,0.5,0.5))

plot_data = lapply(tars, function(x) {
    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
    tmp = prep_plot_data(dd$yield_lb * 1.12, dd$yrs_in_trial)
    tmp$yrs_in_trial = as.numeric(row.names(tmp))
    tmp
    })

invisible(sapply(1:2, function(i){
    plot(mean ~ yrs_in_trial,
         data = plot_data[[i]],
         xlim = c(-6, 36), ylim = c(6500,14500),
         pch = 16, main = gsub("M","M-", tars[i]),
         ylab = "",
         yaxt = "n",
         xlab = "Years after release")
    segments(x0 = plot_data[[i]]$yrs_in_trial,
             y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
             y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
    if(i == 1){
        axis(2)
        mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }
    abline(lm(yield_lb*1.12 ~ yrs_in_trial,
              data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
           lty = 1, col = "grey50")
}))

@@

There is evidence of a small decrease in yield over time once a cultivar is
released. This complicates comparisons between new entries and checks,
since the check will decrease over time. This decrease will manifest
as an apparent larger yield advantage of the new entry over the check.
There is evidence of a small increase in yield with each new released
cultivar. Therefore, it appears that successive cultivar releases are
not merely recovering to the old benchmark but are surpassing
it. However, progress is slower than previously estimated
(\Sexpr{round(fixef(release_fit)[2] * 1000, 1)} lbs
acre\textsuperscript{-1} year\textsuperscript{-1} increase verses a
previous estimate of 54 lbs acre\textsuperscript{-1}
year\textsuperscript{-1}).

The practical impact of these results is: Breeders need to be aware of
the potential for the yield of checks to decrease over time. Due to
this, comparisons to older checks might give a false sense of
improvement. In order to have greater improvements in yield, the
target yield improvement for a new cultivar needs to be higher to
offset the impact of yield decrease once a cultivar is
released. Breeding is essential to maintaining yield in rice. Yield is
not stable and without continuous effort, yield potential will
decrease over time.

\newpage
\section{References}\label{references}
